The Constant-Q Transform Spectral Envelope Coefficients: A Timbre Feature Designed for Music

### 2021-06-29 IEEE SP Magazine/06-2021 - Draft/article.tex
Worked more on the solution section.

\IEEEPARstart{T}{imbre} is the attribute of sound which makes, for example, two musical instruments playing the same note sound different. It is generally associated with the spectral (but also temporal) envelope and is typically assumed to be independent from the pitch (but also the loudness) of the sound \cite{moore2004}. In this article, we will show how to design a simple but functional pitch-independent timbre feature which is well-adapted to musical data, by deriving it from the constant-Q transform (CQT) \cite{brown1991, brown1992}, a log-frequency transform which aligns with the Western music scale. We will show how to decompose the CQT-spectrum into an energy-normalized pitch component and a pitch-independent spectral envelope, the latter from which we will extract a number of timbral coefficients. We will then evaluate the discriminative power of these CQT-spectral envelope coefficients (CQT-SEC) on the NSynth dataset \cite{engel2017}, a large-scale dataset of musical notes which is publicly available, comparing them with the mel-frequency cepstral coefficients (MFCCs) \cite{mermelstein1976}, features originally designed for speech recognition but commonly used to characterize timbre in music.

@@

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one simple characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound, which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when they need one simple timbre feature \cite{mueller2007}. While the MFCCs were shown to be helpful in some MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily well-adapted to musical data. In particular, they are derived through an old process which makes use of the mel scale, a perceptual scale experimentally designed 85 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but generally in terms of implicit embeddings which are tied to a specific trained model \cite{engel2017}, \cite{pons2017}, and not necessarily as explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to feature to characterize musical timbre by MIR practitioners.

@@

We propose here the CQT-SECs, a novel timbre feature which is well-adapted to musical data, pitch-independent, simple to compute, interpretable, and functional. We will show how to derive it from the CQT, a frequency transform with a logarithmic resolution which matches the notes of Western music scale \cite{brown1991, brown1992}, by first decomposing the CQT log-spectrum into a pitch-independent spectral envelope and an energy-normalized pitch component, and then extracting a number of timbral coefficients from the spectral envelope.

@@

Figure \ref{fig:deconvolution} shows an example of deconvolution of a CQT-spectrogram into its envelope and pitch component. The CQT-spectogram was computed from an audio signal created by concatenating 12 4-second notes from an acoustic bass playing from C1 (32.70 Hz) to B1 (61.74 Hz) in ascending order. The notes come from the NSynth dataset \cite{engel2017} and correspond to instrument id \texttt{bass\_acoustic\_000}, MIDI numbers \texttt{024} to \texttt{035}, and velocity number \texttt{075}. As we can see, the envelope looks as if the CQT-spectrogram has been normalized in pitch, with all the notes being brought down to the lowest frequency, here C1, while the pitch component looks as if the CQT-spectrogram has been stripped down from all its energy, leaving mostly the fundamental frequencies of the notes. Note that in practice, we use a power CQT-spectrogram and we take the real part of the envelope and pitch component to ensure real values.

\begin{figure*}[htp]
    \centering
    \includegraphics[width=\textwidth]{deconvolution.png}
    \caption{Deconvolution of the CQT-spectrogram (left plot, shown in dB) of 12 acoustic bass notes playing from C1 to B1, into a pitch-independent envelope (middle plot, shown in dB) and an energy-normalized pitch component (right plot, shown in $[0, 1]$).}
    \label{fig:deconvolution}
\end{figure*}

@@

Once the CQT-spectrum has been decomposed into an envelope and a pitch component, we can then assume that most of the pitch information has been removed from the envelope which now mostly contains the timbre information. As shown in the middle plot of Figure \ref{fig:deconvolution}, the envelope can be thought of as a pitch-normalized log-spectrum where the spectral components, in particular the harmonics which contain most of the energy from the musical instrument, have been essentially brought down to the same note level. Given the octave resolution that was used to compute the CQT-spectrum, i.e., the number of bins per octave, we can then easily infer the locations of those harmonics in the envelope. We can subsequently extract these harmonics, or \textit{spectral envelope coefficients}, from the envelope, and therefore obtain a judicious compact feature for characterizing the timbre of the musical instrument. Equation \ref{eq:extraction} shows how to derive the indices of the spectral envelope coefficients given $O_r$, the octave resolution, and $N_c$, the number of desired coefficients, and consequently extract the CQT-SECs from the envelope $E$, with $\log_2(.)$ and $round(.)$ representing the binary logarithm and the round function, respectively.
\begin{equation}
\label{eq:extraction}
\begin{split}
\begin{cases}
i = round(O_r \log_2(k)) \\
\text{CQT-SEC}_k = E(i) \\
\end{cases}
\quad 1 \le k \le N_c
\end{split}
\end{equation}

>>>>

\IEEEPARstart{T}{imbre} is the attribute of sound which makes, for example, two musical instruments playing the same note sound different. It is generally associated with the spectral (but also temporal) envelope and is typically assumed to be independent from the pitch (but also the loudness) of the sound \cite{moore2004}. In this article, we will show how to design a simple but functional pitch-independent timbre feature which is well-adapted to musical data, by deriving it from the constant-Q transform (CQT) \cite{brown1991, brown1992}, a log-frequency transform which matches the equal temperament musical scale. We will show how to decompose the CQT-spectrum into an energy-normalized pitch component and a pitch-independent spectral envelope, the latter from which we will extract a number of timbral coefficients. We will then evaluate the discriminative power of these CQT-spectral envelope coefficients (CQT-SEC) on the NSynth dataset \cite{engel2017}, a large-scale dataset of musical notes which is publicly available, comparing them with the mel-frequency cepstral coefficients (MFCCs) \cite{mermelstein1976}, features originally designed for speech recognition but commonly used to characterize timbre in music.

@@

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one simple characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound, which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when they need one simple timbre feature \cite{mueller2007}. While the MFCCs were shown to be helpful in some MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily well-adapted to musical data. In particular, they are derived through an old process which makes use of the mel scale, a perceptual scale experimentally designed 85 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but generally in terms of implicit embeddings which are tied to a specific trained model \cite{engel2017}, \cite{pons2017}, and not necessarily as explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to feature to characterize timbre by MIR practitioners.

@@

We propose here the CQT-SECs, a novel timbre feature which is well-adapted to musical data, pitch-independent, simple to compute, interpretable, and functional. We will show how to derive it from the CQT, a frequency transform with a logarithmic resolution which matches the notes of the equal temperament, a musical scale typically used in Western music \cite{brown1991, brown1992}. We will first show how to decompose the CQT-spectrum into a pitch-independent spectral envelope and an energy-normalized pitch component, and then extract a number of timbral coefficients from the spectral envelope.

@@

Figure \ref{fig:deconvolution} shows an example of deconvolution of a CQT-spectrogram into its envelope and pitch component. The CQT-spectogram was computed from an audio signal created by concatenating 12 4-second notes of an acoustic bass playing from C1 (32.70 Hz) to B1 (61.74 Hz) in ascending order. The notes come from the NSynth dataset \cite{engel2017} and correspond to instrument id \texttt{bass\_acoustic\_000}, MIDI numbers \texttt{024} to \texttt{035}, and velocity number \texttt{075}. As we can see, the envelope looks as if the CQT-spectrogram has been normalized in pitch, with all the notes being brought down to the lowest frequency (here, corresponding to C1); while the pitch component looks as if the CQT-spectrogram has been stripped down from all its energy, leaving mostly the fundamental frequencies of the notes. Note that, in practice, we use a power CQT-spectrogram (magnitude to the power of 2) and take the real part of the envelope and pitch component to ensure real values. We also note that this deconvolution can potentially be further refined, for example, by zeroing the few negative values in the pitch component to only have values in $[0, 1]$, and using it to re-derive the spectral envelope from the log-spectrum.

\begin{figure*}[htp]
    \centering
    \includegraphics[width=\textwidth]{deconvolution.png}
    \caption{Deconvolution of the CQT-spectrogram (left plot, shown in dB) of 12 acoustic bass notes playing from C1 to B1, into a pitch-independent spectral envelope (middle plot, shown in dB) and an energy-normalized pitch component (right plot, shown in $[0, 1]$).}
    \label{fig:deconvolution}
\end{figure*}

@@

Once the CQT-spectrum has been decomposed into an envelope and a pitch component, we can then assume that most of the pitch information has been removed from the envelope which now mostly contains the timbre information. As shown in the middle plot of Figure \ref{fig:deconvolution}, the envelope can be thought of as a pitch-normalized log-spectrum where the spectral components, in particular, the harmonics which contain most of the energy of the musical instrument, have been essentially brought down to the same note level. Given the octave resolution that was used to compute the CQT-spectrum, i.e., the number of bins per octave, we can then easily infer the locations of those harmonics in the envelope. We can subsequently extract these harmonics, or \textit{spectral envelope coefficients}, from the envelope, and therefore obtain a compact and meaningful feature for characterizing the timbre of the musical instrument. Equation \ref{eq:extraction} shows how to derive the indices of the spectral envelope coefficients given $O_r$, the octave resolution, and $N_c$, the number of desired coefficients, and consequently extract the CQT-SECs from the envelope $E$, with $\log_2(.)$ and $round(.)$ representing the binary logarithm and the round function, respectively.
\begin{equation}
\label{eq:extraction}
\begin{split}
\begin{cases}
i = round(O_r \log_2(k)) \\
\text{CQT-SEC}_k = E(i) \\
\end{cases}
\quad 1 \le k \le N_c
\end{split}
\end{equation}
